---
title: "ESM 244 - Final"
author: "Jamie K. Miller"
date: "February 14, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
# Install new packages
#install.packages("rgdal")
# install.packages("gridExtra")
# install.packages("grid") # package didn't work
# install.packages("shinythemes")
```


```{r packages, include = FALSE}
# load packages

library(tidyverse)
library(janitor) # to clean column names
library(lubridate) # date wrangling. will probably need to separate out counts by month or weeknumber to make visualization easier/nicer. converts dates to dates rather than numerical sequences.
#library(tseries) # for time series data
#library(sf)
#library(tmap)
#library(rgdal) # For transforming UTM corrdinates to lat/long
#library(leaflet)
#library(ggrepel)
#library(ggspatial)
#library(RColorBrewer)
#library(gganimate)
#library(spatstat)
#library(maptools)
#library(sp)
#library(raster)
#library(gridExtra) # these apckages are for turnding graphs into grid plots I think. Didn't work.
#library(grid) # didn't work


```

###Use this to read in data.
```{r, include = FALSE}
# Load all updated data. Let's try this again.

transect <- read_csv("transect_updated.csv") %>% 
  clean_names()
# From this dataset we want Date, Transect block, total male, total female, unknown, and juvs. I added a totals column in excel to double check numbers. Associated math is in next code chunk.

avian <- read_csv("avian_updated.csv")%>% 
  clean_names() %>% 
#  dplyr::select(survey_date, transect_block, species, count)
  dplyr::select(survey_week, transect_block, species, count)
# From this dataset we want date, Transect block, species, and count
# modified data in excel so date is standardized to monday of survey week. this will separate out surveys that span 2 days.

wrack_df <- read_csv("wrack_updated.csv")%>% 
  clean_names()
# wrack values, date, and transect block

lookup_avian <- read_csv("lookup_avian.csv")%>% 
  clean_names() %>% 
  rename(species = actual_species) %>% 
  filter(analyze == 1) # filtered a subset of birds to analyze
# Has 4-letter avian code from AOU species list found here: https://www.pwrc.usgs.gov/bbl/manual/speclist.cfm or species_list.pdf in repo

lookup_blocks <- read_csv("transect_block_coords.csv")%>% 
  clean_names() %>% 
  filter(beach == "NORTH" | beach == "SOUTH" | beach == "PURISIMA")
# transect block names, beach section initial code, region designation, and gps coords

```

###Wrangle SNPL population counts from transect data.
```{r}
# Tally SNPL count from transect data so it can be added to the avian data
# SNPL data is collected be age/sex demographic Need to create totals and filter out all transect blocks that do not have any plovers in them.

snpl <- transect %>%
  dplyr::select(survey_week, transect_block, beach_section_initials, beach, males_individual, females_individual, unknown_individual, pairs, juvenile, total) %>% 
  mutate(total_male = males_individual + pairs) %>% # combine M individ. w/ pairs to get total number of males
  mutate(total_female = females_individual + pairs) %>% # combine F individ. w/ pairs to get total number of females
  mutate(snpl = total_male + total_female + unknown_individual + juvenile) %>% # this should match column 'total'. Looks like it does
  filter(snpl > 0) # filters out blocks where non SNPL were seen

snpl_clean <- snpl %>%
  dplyr::select(survey_week, transect_block, beach_section_initials, beach, snpl) %>% 
  rename(count = snpl) %>% 
  filter(beach == "NORTH" | beach == "PURISIMA" | beach == "SOUTH") 

```

###Wrangle avian data
```{r}
# Need to join avian data with lookup_avian for species name and type info, and lookup_blocks for beach section.

# Join avian species info to avian count data
avian_sp_join <- full_join(avian, lookup_avian, by = "species") %>%  # yes, this worked.
  filter(analyze == 1)

# Clean up transect blocks before joining. 
blocks <- lookup_blocks %>% 
  dplyr::select(transect_block, beach_section_initials, beach, y_proj, x_proj)

# Join blocks to avian
avian_block_join <- full_join(avian_sp_join, blocks, by = "transect_block")

# Clean up columns and filter for only North, Purisima, and South beaches 
avian_clean <- avian_block_join %>% 
  dplyr::select(survey_week, transect_block, y_proj, x_proj, beach_section_initials, beach, species, species_name, species_type, count) %>% 
  filter(beach == "NORTH" | beach == "PURISIMA" | beach == "SOUTH") %>% 
  filter(species != "NA")

```

###Combine avian and snpl data for a complete VAFB birds dataset
```{r}
# Merge test # 1
# Merge snpl_clean with avian_clean into new df
# then all SNPL entries will show NA in species, species_name, and species_type

vafb_birds_combine <- bind_rows(avian_clean, snpl_clean) %>% # yay it worked!
  filter(count > 0) 

# The following three code snippets convert 'NA' values in the species, species_name, and species_type columns to 'SNPL', 'Snowy Plover', and 'Shorebird', repectively. Each column needed to be a character for it to work, according to Google.

vafb_birds_combine$species <- as.character(vafb_birds_combine$species)
vafb_birds_combine$species <- ifelse(is.na(vafb_birds_combine$species),'SNPL',vafb_birds_combine$species)

vafb_birds_combine$species_name <- as.character(vafb_birds_combine$species_name)
vafb_birds_combine$species_name <- ifelse(is.na(vafb_birds_combine$species_name),'Snowy Plover',vafb_birds_combine$species_name)

vafb_birds_combine$species_type <- as.character(vafb_birds_combine$species_type)
vafb_birds_combine$species_type <- ifelse(is.na(vafb_birds_combine$species_type),'Shorebird',vafb_birds_combine$species_type)

# ok, so now we can use this data to make some graphs
# will probably need to summarize at some point

```

###Use lubridate() functions to convert to times/dates
```{r date_time}
vafb_birds_combine$survey_week <- mdy(vafb_birds_combine$survey_week)

```

###For the time series plots, group by beach section. Use these dataframes for making graphs!!!
```{r}
# Sum counts for each beach section. can then plot max birds per beach section over time. Keep transect block separations for map.


birds_by_sector <- vafb_birds_combine %>% 
  group_by(survey_week, beach_section_initials, beach, species, species_type) %>% 
  summarize(total = sum(count))


birds_by_region <- vafb_birds_combine %>% 
  group_by(survey_week, beach, species_type) %>%  # remove species
  summarize(total = sum(count))

write_csv(birds_by_region, "birds_by_region.csv")

```


```{r, fig.width=30, include = FALSE}

birds_by_region %>% 
#  filter(species == "SNPL") %>% 
  filter(beach == "PURISIMA") %>% 
  ggplot() +
  geom_point(aes(x = survey_week, y = total)) +
  geom_line(aes(x = survey_week, y = total)) #+
#  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))


```


###Wrangle the wrack data to make it useful
```{r, include = FALSE}
wrack_simple <- wrack_df %>% 
  dplyr::select(survey_week, transect_block, wrack) %>%
  filter(wrack != -9) 

#wrack_simple$survey_week <- mdy(wrack_simple$survey_week) # This line of code just messes things up.

# Attach transect block info to wrack data so we can select it in dropdowns
wrack_beach <- full_join(wrack_simple, lookup_blocks, by = "transect_block") %>% 
  dplyr::select(survey_week, transect_block, wrack, beach_section_initials, beach) %>% 
  filter(beach == "NORTH" | beach == "SOUTH" | beach == "PURISIMA")

# Not certain this is necessary, but it yields the mean wrack values for each beach sector per survey day
wrack_mean_sector <- wrack_beach %>% 
  group_by(survey_week, beach_section_initials) %>% 
  summarize(mean = mean(wrack))



```




###Tab 1 Graph
```{r, fig.width=30, include = FALSE}


#  filter(species == "SNPL") %>% 
#  filter(beach == "PURISIMA") %>% 
  ggplot(birds_by_region) +
  geom_point(aes(x = survey_week, y = total)) +
  geom_line(aes(x = survey_week, y = total)) +
  labs(x = "Date", y = "Number Birds Observed", title = "VAFB Birds") +
  facet_wrap(~species_type, scale = "free")

# Figure out how to make this selectable in widgets. 

# Widget 1: Dropdown/radio button selects species or species type. 
# Widget 2: Slider scale selects date. Range of dates.
# Widget 3: Dropdown/radio button selects beach sector (MIN, SHN, SHN, etc.) or region (North/Purisima/South).

# Result: Graph of species or species type population time series. Grouped by Beach Sector at smallest.
  
# Incorporate gganimate!

```

###Tab 2 Map
```{r}
# Map prep code

# Read in couties shapefile for a map background
ca_counties <- read_sf(".", layer = "california_county_shape_file") #%>% 
  st_transform(crs = 4326) # I guess it already is. R didn't like this snippet of code.

# Filter to Santa Barbara County. We will be soomed in enough that we don't need other background shapefiles.
sb_county <- ca_counties %>% 
  filter(NAME == "Santa Barbara") #%>% 
 # Double check crs for WGS84 UTM ZOne 10S datum that transect block points are in. Or just transform them all. I'm pretty sure they are all transformed.

# Simplfy lookup_blocks so only transect block, and UTM are shown. Need to convert UTM to lat/long
block_gps <- lookup_blocks %>% 
  dplyr::select(x_proj, y_proj) %>% 
  filter(x_proj != "NA")

rownames(block_gps) <- lookup_blocks$transect_block # Didn't work. I think that's ok.

#Transform UTM to lat/long
block_utm <- SpatialPoints(block_gps, proj4string=CRS("+proj=utm +zone=10S +datum=WGS84"))

block_latlong <- spTransform(block_utm, CRS("+proj=longlat +datum=WGS84"))

# Coerce back to dataframe so ggplot knows what to do with it.
block_gps2 <- as.data.frame(block_latlong) 

plot(block_gps2) # Just checking that it works. It does.
    

# Set extent to y_proj 34.87 to 34.60, x_proj -120.65 to -120.59
bounds <- as(extent(-120.65, -120.59, 34.60, 34.87), 'SpatialPolygons')
crs(bounds) <- crs(sb_county)
sb_crop <- raster::crop(sb_county, bounds) # only worked once. but did work.

sb_crop2 <- st_crop(sb_county, c(xmin = -120.65, xmax = -120.59, ymin = 34.63, ymax = 34.87)) # This one works and is more streamlined.

```

```{r, fig.height = 10, fig.width = 2}

# Trying out map code
#ggplot(sb_crop2) +
#  geom_sf(aes(fill = STATE),
#          fill = "darkgreen",
 #         color = "NA",
  #        show.legend = FALSE) +
  #geom_point(data = block_gps2,
#             aes(x = x_proj, y = y_proj),
 #            color = "black") +
  #theme(panel.background = element_rect(fill = 'blue', colour = 'NA')) + # can change these to better colors
#  coord_sf(datum = NA) +
 # labs(x = "", 
  #     y = "")
  
# ok now we just need to get bird data to show up on here. 

# I think what I need to do for bird data is attach bird data to block_gps2 as attributes. Or use tmap.

#map_wrack <- tm_shape(sb_crop2) +
  
######################  
  
# USE vafb_birds_combine. it has gps attached now.

#########################

# Widget 1: Dropdown/radio button selects species or species type. 
# Widget 2: Slider scale selects date. 
# Widget 3: Dropdown/radio button selects region (North/Purisima/South). Might be more informative to just keep at at a North/Purisima vs. South Option.

# Result: Size of dot reflects count. Grouped by transect block at smallest.
# Interactive bubble plot.


# Use avian_block_join table to make bird pop geom_point graph where size = count. perhaps color = species if select multiple species.

```



###Tab 3 Graphs with wrack data
```{r}

wrack_shore_gull <- birds_by_sector %>% 
  filter(species == "SNPL" | species == "SAND" | species_type == "Gull") %>% 
  mutate(sp_2 = case_when(
    species == "SNPL" ~ "SNPL",
    species == "SAND" ~ "SAND",
    species == "CAGU" ~ "Gull",
    species == "WEGU" ~ "Gull",
    species == "HEEG" ~ "Gull"
  ))

```

```{r, include=FALSE}
wrack_plot <- ggplot(wrack_mean_sector, aes(x = survey_week, y = mean)) +
  geom_col() 
wrack_plot

shore_gull_plot <- ggplot(wrack_shore_gull, aes(x = survey_week, y = total)) +  
  geom_line(aes(color = sp_2)) 
shore_gull_plot


# Tried this and it didn't work.
# http://steffi.ca/thinkR/tutorials/twoplots/
#grid.arrange(wrack_plot,shore_gull_plot, heights = c(1/5, 4/5))

# Widget 1: Dropdown/radio button selects species or species type. 
# Widget 2: Slider scale selects date. 
# Widget 3: Dropdown/radio button selects beach sector (MIN, SHN, SHN, etc.) or region (North/Purisima/South).

# Result: Graph of species or species type population time series. Limit this to species::SNPL, species::SAND, and species_type::gulls. Grouped by Beach Sector at smallest.

# the only gulls used will be the three species from lookup_avian -> analyze = 1.

# Incorporate gganimate!


# wrack data wuse wrack_mean_sector
# bird data use filtered birds_by_sector

# on single plot x anix is time, one y axis is mean daily wrack data. sepaparated by beach sector
# second y axis is bird population : SNPL, SAND, gull.

# bar graph overlayed by line plt. bar plot is mean wrack.
# line plot is bird. if select all 3, then separate by color.


```


```{r}
# in this new code snippet I anm going to join wrack_shore_gull and wrack_mean_sector by date and beach_section_initials.

# first, get dates the same in both data frames
wrack_mean_sector$survey_week <- mdy(wrack_mean_sector$survey_week)

write_csv(wrack_mean_sector, "wrack_mean.csv")

wrack_avian_join <- full_join(wrack_shore_gull, wrack_mean_sector, by = c("survey_week", "beach_section_initials")) 

wrack_avian_df <- wrack_avian_join %>% 
  dplyr::select(survey_week, beach_section_initials, sp_2, total, mean) %>% 
  rename(species_2 = sp_2, mean_wrack = mean) %>% 
  filter(mean_wrack != "NA") # removed rows where wrack = NA, meaning no wrack data were collected

write_csv(wrack_avian_df, "wrack_avian_df.csv") 

```



###What we need to do:
- Background/Introductory Tab 
     - Can choose from jpegs in GitHub folder for informative base map.
     - Will add a snowy plover photo. 
- Create app shell
- Tab 1 line graphs
- Tab 2 map with population represented as dots
- Tab 3 line graphs with wrack data. keep it simple and use just SNPL, SAND, and species type::Gulls.
- Figure out how to attach our graphs to the app. How do the controls interact with the graphs?




##### Data Tab 1 Time and Species Count Data
Widget 1: Input: Time by year (2012 - 2018)
Type: Slider, which will allow app users to examine progression of data on a temporal scale. 

Widget 2
Input: Location of observation site
Type: Radio button, which will allow app users to switch between observation sites. 

Widget 3
Input: Bird species
Type: Checkboxes or multiple select box, which will allow app users to compare multiple species on a single output. 

App users will be able to use the time slider to scroll through temporal data and species count data, which will generate a line graph. 


##### Data Tab 2 Abundance and Study Map

Widget 1
Input: Time by year (2012 - 2018)
Type: Slider, which will allow app users to examine progression of data on a temporal scale. 


Widget 2
Input: Location of observation site
Type: Radio button, which will allow app users to switch between observation sites. 


Widget 3
Input: Bird species
Type: Checkboxes or multiple select box, which will allow app users to compare multiple species on a single output. 

App users will be able to view changes in species density over time by selecting one or more species, and projecting species density onto a map of the project area. 



##### Data Tab 3 Wrack Data and SNPL/SAND/gull graphs

Widget 1
Input: Time by year (2012 - 2018)
Type: Slider, which will allow app users to examine progression of data on a temporal scale. 


Widget 2
Input: Location of observation site
Type: Radio button, which will allow app users to switch between observation sites. 


Widget 3
Input: Shorebird species
Type: Checkboxes or multiple select box, which will allow app users to compare multiple species on a single output. 

App users will be able to use the time slider to scroll through wrack index data and shorebird count data, which will generate a line graph. 


##### Below is just a copy/paste of our summary that we turned in. Not formatted. Just information.
App Summary
This dataset was collected for Vandenberg Air Force Base (VAFB) by Point Blue Conservation Science, a nonprofit based in central California, as part of ongoing beach ecosystem and snowy plover conservation projects. Our app will explore seasonal and annual fluctuations in shorebird, gull, and raptor abundance, and changes in beach habitat characteristics at VAFB.

Data
Since March 2012, weekly transect surveys were conducted at VAFB beaches. Each beach sector was divided into “transect blocks” approximately 100-300 meters in length along the coastal strand. Within each transect block, counts were taken of the number of snowy plovers, age, sex, flock size, presence of paired individuals, and presence of broods. Additionally, the number and species of shorebirds, seabird, or raptors utilizing the habitat was recorded, and the amount of wrack present on each block was scored (Robinette et al. 2017).

Thousands of data points from bird count transects have been collected by field biologists since 2012. The data are not in tidy format. Data have not been recorded for each bird observed-- bird counts are aggregated and reported at the end of each transect and/or field survey. We will be able to convert the data to tidy format if necessary. 

We will use the following variables in our app, which include data from weekly and semi-weekly field surveys:  
Date of survey
Location of survey site
Bird type: Category (shorebird, gull, or raptor), Species
Species/Category abundance: # of observed snowy plovers (based on age & sex), shorebirds, gulls, or raptors 
Wrack Index: category assigned to abundance of fresh wrack (surf-cast kelp) on the beach (e.g. a rating of 1 would be the least amount of wrack, and 5 is the highest amount of wrack)


Widgets
We will include the following widgets in our app: 
Widget 1
Input: Time by year (2012 - 2018)
Type: Slider, which will allow app users to examine progression of data on a temporal scale. 
Widget 2
Input: Location of observation site
Type: Radio button, which will allow app users to change location scale and switch between beach sectors. 
Widget 3
Input: Bird species
Type: Checkboxes or multiple select box, which will allow app users to compare multiple species or an entire category on a single output. 

Outputs
We will include the following three final outputs that will be produced based on users’ widget selections: 

Output 1
 App users will be able to use the time slider to scroll through temporal data and species count data, which will generate a line graph. 

Output 2
App users will be able to view changes in species density over time by selecting one or more species, and projecting species density onto a map of the project area.
Output 3
App users will be able to use the time slider to scroll through wrack index data and shorebird count data, which will generate a line graph. 

Literature Cited
Robinette, D.P, J.K. Miller, and J. Howar. 2017. Monitoring and Management of the Endangered California Least Tern and the Threatened Western Snowy Plover at Vandenberg Air Force Base, 2017. Unpublished Report, Point Blue Conservation Science, Petaluma, CA.


##### Challenges
- Data wrangling a problematic data set
- really allowed me to identify data inconsistencies

