---
title: "ESM 244 - Final - old code"
author: "Jamie Miller"
date: "March 6, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# This document is old code tjhat we don't need anymore.
The following code chunks were tried and either didn't work or contain snippets that are no longer valid. Lots of data wrangling and visualizations and trouble-shooting. They are cluttering up my workspace. Keeping them in case something in them is useful or informative. Doubt it, but here it is. -JKM 3/6/19


###This code chunk is old - do not use.
```{r, include = FALSE}
# Load data

#transect <- read_csv("transect.csv") %>% 
#  clean_names()

# From this dataset we want Date, Transect block, total male, total female, unknown, juvs, and wrack

#avian <- read_csv("avian.csv")%>% 
#  clean_names()

# From this dataset we want date, Transect block, species, and count

```

###This code chunk is old - do not use.
```{r, include = FALSE}
# Copied this code chunk below. Will modify for updated datasets. This code chunk only worked woth the original data.
# Will reuse dataframe names.

#snpl <- transect %>%
#  select(survey_date, transect_block, total_male, total_female, unknown_individual, juvenile) %>% 
#  mutate(snpl = total_male + total_female + unknown_individual + juvenile) %>% 
#  filter(snpl > 0)

#snpl_clean <- snpl %>%
#  select(survey_date, transect_block, snpl) %>% 
#  rename(count = snpl)

#wrack <- transect %>%
#  select(survey_date, transect_block, wrack)

#avian_clean <- avian %>% 
#  select(survey_date, transect_block, species, species_name, species_type, count)

```





###Visualization attempts
```{r include = FALSE}

#ggplot(vafb_birds_combine) +
 # geom_point(aes(x = survey_week, y = count, color = species))

# this looks awful

```

```{r, include = FALSE}

#ggplot(vafb_birds_combine) +
#  geom_point(aes(x = survey_week, y = count, color = species)) +
#  facet_wrap(~species_type, scale = "free") +
#  theme(legend.position = "none")

# better. Using species type as a dropdown for the app will be useful.


```


```{r, include = FALSE}
#ggplot(vafb_birds_combine) +
#  geom_point(aes(x = survey_date, y = count)) +
#  facet_wrap(~species_type, scale = "free") +
#  theme(legend.position = "none")

# better. Using species type as a dropdown for the app will be useful.
# why are there 1000 terns? that doesn't seem right. Checked the data and it appears entered accurately.

```


```{r, include = FALSE}
# Look at just shorebirds for a moment
#vafb_birds_combine %>% 
#  filter(species_type == "Shorebird") %>% 
#  ggplot() +
#  geom_point(aes(x = survey_date, y = count)) +
#  facet_wrap(~species, scale = "free") 

# better. Using species type as a dropdown for the app will be useful.


```

```{r, include = FALSE}
# Look at just SNPL and SAND for a moment
#vafb_birds_combine %>% 
 # filter(species == "SNPL" | species == "SAND") %>% 
 # ggplot() +
#  geom_point(aes(x = survey_date, y = count)) +
#  facet_wrap(~species, scale = "free") 

# I feel like something wierd is going on with the dates. Wrangle that next. I don't think it is sorted chronologically.
# Maybe by combining with wrack data it will sort out.
```

```{r, include = FALSE}
# Look at just SNPL separated by beach sector. Maybe this will help sort out dates.
#vafb_birds_combine %>% 
#  filter(species == "SNPL") %>% 
#  ggplot() +
#  geom_point(aes(x = survey_date, y = count)) +
#  facet_wrap(~beach, scale = "free") 

# Why is Jalama still here? I removed it earlier, yes? - ok fixed it in snpl_clean dataframe.
```


```{r, fig.width=30, include = FALSE}
# Look at just SNPL on North Beaches. Maybe this will help sort out dates.
vafb_birds_combine %>% 
  filter(species == "SNPL") %>% 
  filter(beach == "NORTH") %>% 
  ggplot() +
  geom_point(aes(x = survey_week, y = count)) 


# Look at just SNPL on Purisima Beaches. Maybe this will help sort out dates.
vafb_birds_combine %>% 
  filter(species == "SNPL") %>% 
  filter(beach == "PURISIMA") %>% 
  ggplot() +
  geom_point(aes(x = survey_week, y = count, color = beach_section_initials)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

# ok, for the graphs we need a count for each beach sector. Think about how fine scale we want to split this up.
# also, dates are sorted by month/date/year numerically. Not chronologically. Hopefully lubridate can help with this.
# Sorted out dates with lubridate. - fixed it in prior code chunk
# Also, maybe can try re-filtering so we include blocks with 0 counts of birds.

```





