---
title: "ESM 244 - Final Markdown"
author: "Jamie Miller"
date: "March 11, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
# load packages

library(tidyverse)
library(janitor) # to clean column names
library(lubridate) # date wrangling. will probably need to separate out counts by month or weeknumber to make visualization easier/nicer. converts dates to dates rather than numerical sequences.
#library(tseries) # for time series data

#Mapping packages
library(sf)
library(leaflet)
library(sp)

```

###Use this to read in data.
```{r, include = FALSE}
# Load all updated data. Let's try this again.

transect <- read_csv("transect_updated.csv") %>% 
  clean_names()
# From this dataset we want Date, Transect block, total male, total female, unknown, and juvs. I added a totals column in excel to double check numbers. Associated math is in next code chunk.

avian <- read_csv("avian_updated.csv")%>% 
  clean_names() %>% 
#  dplyr::select(survey_date, transect_block, species, count)
  dplyr::select(survey_week, transect_block, species, count)
# From this dataset we want date, Transect block, species, and count
# modified data in excel so date is standardized to monday of survey week. this will separate out surveys that span 2 days.

wrack_df <- read_csv("wrack_updated.csv")%>% 
  clean_names()
# wrack values, date, and transect block

lookup_avian <- read_csv("lookup_avian.csv")%>% 
  clean_names() %>% 
  rename(species = actual_species) 
# Has 4-letter avian code from AOU species list found here: https://www.pwrc.usgs.gov/bbl/manual/speclist.cfm or species_list.pdf in repo

lookup_blocks <- read_csv("transect_block_coords.csv")%>% 
  clean_names() %>% 
  filter(beach == "NORTH" | beach == "SOUTH" | beach == "PURISIMA")
# transect block names, beach section initial code, region designation, and gps coords

```

###Wrangle SNPL population counts from transect data.
```{r}
# Tally SNPL count from transect data so it can be added to the avian data
# SNPL data is collected be age/sex demographic Need to create totals and filter out all transect blocks that do not have any plovers in them.

snpl <- transect %>%
  dplyr::select(survey_week, transect_block, beach_section_initials, beach, males_individual, females_individual, unknown_individual, pairs, juvenile, total) %>% 
  mutate(total_male = males_individual + pairs) %>% # combine M individ. w/ pairs to get total number of males
  mutate(total_female = females_individual + pairs) %>% # combine F individ. w/ pairs to get total number of females
  mutate(snpl = total_male + total_female + unknown_individual + juvenile) %>% # this should match column 'total'. Looks like it does
  filter(snpl > 0) # filters out blocks where non SNPL were seen

snpl_clean <- snpl %>%
  dplyr::select(survey_week, transect_block, beach_section_initials, beach, snpl) %>% 
  rename(count = snpl) %>% 
  filter(beach == "NORTH" | beach == "PURISIMA" | beach == "SOUTH") 

```

###Wrangle avian data
```{r}
# Need to join avian data with lookup_avian for species name and type info, and lookup_blocks for beach section.

# Join avian species info to avian count data
avian_sp_join <- full_join(avian, lookup_avian, by = "species") #%>%  # yes, this worked.
#  filter(analyze == 1)

# Clean up transect blocks before joining. 
blocks <- lookup_blocks %>% 
  dplyr::select(transect_block, beach_section_initials, beach, y_proj, x_proj)

# Join blocks to avian
avian_block_join <- full_join(avian_sp_join, blocks, by = "transect_block")

# Clean up columns and filter for only North, Purisima, and South beaches 
avian_clean <- avian_block_join %>% 
  dplyr::select(survey_week, transect_block, y_proj, x_proj, beach_section_initials, beach, species, species_name, species_type, count) %>% 
  filter(beach == "NORTH" | beach == "PURISIMA" | beach == "SOUTH") %>% 
  filter(species_type == "Gull" | species_type == "Raptor" | species_type == "Shorebird" | species_type == "Tern") %>% # filtered a subset of birds to analyze%>% 
  filter(species != "NA")

```

###Combine avian and snpl data for a complete VAFB birds dataset
```{r}
# Merge test # 1
# Merge snpl_clean with avian_clean into new df
# then all SNPL entries will show NA in species, species_name, and species_type

vafb_birds_combine <- bind_rows(avian_clean, snpl_clean) %>% # yay it worked!
  filter(count > 0) 

# The following three code snippets convert 'NA' values in the species, species_name, and species_type columns to 'SNPL', 'Snowy Plover', and 'Shorebird', repectively. Each column needed to be a character for it to work, according to Google.

vafb_birds_combine$species <- as.character(vafb_birds_combine$species)
vafb_birds_combine$species <- ifelse(is.na(vafb_birds_combine$species),'SNPL',vafb_birds_combine$species)

vafb_birds_combine$species_name <- as.character(vafb_birds_combine$species_name)
vafb_birds_combine$species_name <- ifelse(is.na(vafb_birds_combine$species_name),'Snowy Plover',vafb_birds_combine$species_name)

vafb_birds_combine$species_type <- as.character(vafb_birds_combine$species_type)
vafb_birds_combine$species_type <- ifelse(is.na(vafb_birds_combine$species_type),'Shorebird',vafb_birds_combine$species_type)

# ok, so now we can use this data to make some graphs
# will probably need to summarize at some point

```

###For the time series plots, group by beach section. Use these dataframes for making graphs!!!
```{r}
# Sum counts for each beach section. can then plot max birds per beach section over time. Keep transect block separations for map.


birds_by_sector <- vafb_birds_combine %>% 
  group_by(survey_week, beach_section_initials, beach, species, species_type) %>% 
  summarize(total = sum(count))

birds_by_sector$survey_week <- mdy(birds_by_sector$survey_week) #Use lubridate() functions to convert to times/dates

write_csv(birds_by_sector, "birds_by_sector.csv")


birds_by_region <- vafb_birds_combine %>% 
  group_by(survey_week, beach, species_type) %>%  # remove species
  summarize(total = sum(count))

birds_by_region$survey_week <- mdy(birds_by_region$survey_week) #Use lubridate() functions to convert to times/dates

write_csv(birds_by_region, "birds_by_region.csv")

```

###Wrangle the wrack data to make it useful
```{r, include = FALSE}
wrack_simple <- wrack_df %>% 
  dplyr::select(survey_week, transect_block, wrack) %>%
  filter(wrack != -9) 

#wrack_simple$survey_week <- mdy(wrack_simple$survey_week) # This line of code just messes things up.

# Attach transect block info to wrack data so we can select it in dropdowns
wrack_beach <- full_join(wrack_simple, lookup_blocks, by = "transect_block") %>% 
  dplyr::select(survey_week, transect_block, wrack, beach_section_initials, beach) %>% 
  filter(beach == "NORTH" | beach == "SOUTH" | beach == "PURISIMA")

# Not certain this is necessary, but it yields the mean wrack values for each beach sector per survey day
wrack_mean_sector <- wrack_beach %>% 
  group_by(survey_week, beach_section_initials) %>% 
  summarize(mean = mean(wrack)) %>% 
  filter(mean != "NA")

# in this new code snippet I anm going to join wrack_shore_gull and wrack_mean_sector by date and beach_section_initials.

# first, get dates the same in both data frames
wrack_mean_sector$survey_week <- mdy(wrack_mean_sector$survey_week) # lubridate for wrack

write_csv(wrack_mean_sector, "wrack_mean.csv")

```

###Tab 3 Graphs with wrack data
```{r}

wrack_shore_gull <- birds_by_sector %>% 
  mutate(sp_2 = case_when(
    species == "SNPL" ~ "SNPL",
    species == "SAND" ~ "SAND",
    species == "CAGU" ~ "Gull",
    species == "WEGU" ~ "Gull",
    species == "HEEG" ~ "Gull",
    species == "gull" ~ "Gull")) %>%
  filter(species == "SNPL" | species == "SAND" | species_type == "Gull")
  

# below I am going to join wrack_shore_gull and wrack_mean_sector by date and beach_section_initials.

wrack_shore_gull$survey_week <- mdy(wrack_shore_gull$survey_week) # lubridate for wrack_shore_gull so it can be combined with wrack_mean_sector

wrack_avian_join <- full_join(wrack_shore_gull, wrack_mean_sector, by = c("survey_week", "beach_section_initials")) 

wrack_avian_df <- wrack_avian_join %>% 
  dplyr::select(survey_week, beach_section_initials, sp_2, total, mean) %>% 
  rename(species_2 = sp_2, mean_wrack = mean) %>% 
  filter(mean_wrack != "NA") # removed rows where wrack = NA, meaning no wrack data were collected

write_csv(wrack_avian_df, "wrack_avian_df.csv") 

```

#DO NOT CHANGE ANY OF THE ABOVE CODE!!!!
#ANYTHING NEW GOES BELOW THIS LINE. ON PENALTY OF SADNESS.

# This is our attempt to make a map.
```{r}
# comine vafb_birds_combine with block to make sure SNPL has GPS coords without messing up our original dataframe.

# simplify blocks a bit
blocks_2 <- blocks %>% 
  dplyr::select(transect_block, y_proj, x_proj)

avian_block_map <- vafb_birds_combine %>% 
  dplyr::select(survey_week, transect_block, y_proj, x_proj, species, count)

#avian_block_map$y_proj <- as.numeric(avian_block_map$y_proj)
#avian_block_map$y_proj <- ifelse(is.na(avian_block_map$y_proj), full_join(avian_block_map, blocks_2, by = "transect_block"), avian_block_map$y_proj)
# this did not work. horribly so.

write_csv(blocks_2, "blocks_2.csv")

write_csv(avian_block_map, "avian_block_map.csv")

# did the y_proj/x_proj combine step in excel with vlookup 
# use avian_block_map_2.csv

```

```{r}

# Simplfy lookup_blocks so only transect block, and UTM are shown. Need to convert UTM to lat/long
#block_gps <- lookup_blocks %>% 
#  dplyr::select(transect_block, x_proj, y_proj) %>% 
#  filter(x_proj != "NA")

#rownames(block_gps) <- lookup_blocks$transect_block # Didn't work. I think that's ok.


#Transform UTM to lat/long

#block_utm <- SpatialPoints(block_gps, proj4string=CRS("+proj=utm +zone=10S +datum=WGS84"))

#block_latlong <- spTransform(block_utm, CRS("+proj=longlat +datum=WGS84"))

# Coerce back to dataframe so ggplot knows what to do with it.
#block_gps2 <- as.data.frame(block_latlong) 


```

```{r}
# Now can we join blocks_gps2 to avian_block_map_2?

#avian_block_map_2 <- read_csv("avian_block_map_2.csv")

#avian_block_gps <- full_join()


```

#Regression test
```{r}

wrack_avian_df$species_2 <- fct_relevel(wrack_avian_df$species_2, "SAND")

wrack_avian_lm <- lm(total ~ mean_wrack + species_2, data = wrack_avian_df)
wrack_avian_lm

summary(wrack_avian_lm)

```


